Bayesian exponential random graph modeling of KWL Burial network

### 2020-09-29 analysis/paper/paper.Rmd
edit ERGMs section

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical model for expressing properties of network structures with a focus directly on possible ties between nodes by viewing them as random variables [@Robins2007]. In traditional statistical model for network, each predictor variable for forming a tie between two nodes is measure separately by the presence of a tie without considering interactions among ties [@Morris2008]. By contrast, statistical models assumes social networks as dependent variables that can represent the network dependencies between tie variables [@Snijders2011]. In a network generated by ERGMs, the formation of a tie is influenced by the presence or absence of other ties or attributes of nodes/ties. Predictor variables, called network statistics, such as reciprocity, transitivity, homophily, are direct functions of the ties themselves where the probability of occurrence of the configuration of ties can be specified and hypothesized [@Morris2008; @Robins2007]. The advantage of networks specification under assumptions and contexts can be useful to explain archaeological network in a given time. For example, @Amati2019 demonstrate the use of ERGMs to reconstruct networks consist of a set of sites in the Caribbean during the period AD 100–400. They found the structure of network can be efficiently explained by multiple interdependent mechanisms instead of hypotheses of distance and cultural homophily exclusively. However, they also point out some limitations of application such as static outcome and sensitivity to missing data. Another limitation is that ERGMs are difficult to compute because their normalizing constant depending on model parameters is intractable. With the development of Markov chain Monte Carlo (MCMC) algorithms that produce approximate maximum likelihood estimators and graph statistics for further specification of transitivity and heterogeneity, ERGMs can deal with complicated dependence patterns observed at one moment in time [@Snijders2006]. A Bayesian framework allows for parameter inference using Markov chain Monte Carlo (MCMC) strategies which avoid the need for computationally intensive calculations of the normalizing constants [@Caimo2011].

@@

We analyzed social networks using the R programming language with bergm package for Bayesian exponential random graph models [@Caimo2014], which built upon ergm for model specification [@Hunter2008; @Morris2008; @Handcock2008]. Although exponential random graph models can describe a network by capturing the dependency structure, it is difficult to estimate model parameters and interpret the result due to the intractable normalizing constant and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is an important step to make probability distribution to integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. Degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. To overcome these limitations in ERGM, Bayesian model applies Markov chain Monte Carlo (MCMC) simulation with the approximate exchange algorithm to estimate the double intractable posteriors [@Caimo2011]. This alleviate above computational problems. By fitting an ERMG with MCMC algorithm, a Bayesian approach provides posterior probability that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability enables us to estimate the effect of ERGM parameter by observing the posterior mean and 95% credible intervals, which can replace p-value for accessing a null hypothesis [@Caimo2017bayesian]. The criteria is that the odds ratio grater than one means positive effect of ERGM parameter, while odds ratio less than one represents negative effect.

>>>>

# Exponential random graph models

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical model for expressing properties of network structures with a focus directly on possible ties between nodes by viewing them as random variables [@Robins2007]. In traditional statistical model for network, each predictor variable for forming a tie between two nodes is measure separately by the presence of a tie without considering interactions among ties [@Morris2008]. By contrast, statistical models assumes social networks as dependent variables and allows a direct modeling for the network dependencies [@Snijders2011]. In a network generated by ERGMs, the formation of a tie is influenced by the presence or absence of other ties or attributes of nodes/ties. One can model network dependencies according to a common experience or research questions to represent relationships between actors through network variables. For example, reciprocity is a variable describing two actors' relationship, and transitivity describes two actors' relationship via a shared third actor. Those predictor variables, also called network statistics, are direct functions of the ties themselves where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. The advantage of networks specification under assumptions and contexts is useful to explain archaeological network in a given time. For instance, @Amati2019 demonstrated the use of ERGMs to reconstruct three different network for 15 sites in the Caribbean between AD 100 to 400 to explore the interaction mechanisms. Those models test the hypotheses for the relations between sites in terms of proximity, inter-cultural, and exchange network to explain the presence of hub sites. They found that the observed presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of the hypothesis of proximity or cultural homophily exclusively. However, they also point out some limitations of application, such as static outcome and sensitivity to missing data. In addition, it could be difficult to work on ERGMs because their normalizing constant depending on model parameters is intractable. A Bayesian framework allows for parameter inference using Markov chain Monte Carlo (MCMC) algorithm which avoid the need for computationally intensive calculations of the normalizing constants [@Caimo2011]. MCMC algorithms can produce approximate maximum likelihood estimators and graph statistics for further specification of transitivity and heterogeneity, this enables ERGMs to deal with complicated dependence patterns [@Snijders2006].

@@

We analyzed social networks using the R programming language with bergm package for Bayesian exponential random graph models [@Caimo2014], which built upon ergm package for simulation and model specification [@Hunter2008; @Morris2008; @Handcock2008]. Although exponential random graph models can describe a network by capturing the dependency structure, it is difficult to estimate model parameters and interpret the result due to the intractable normalizing constant and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is an important step to make probability distribution to integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. Degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. To overcome these limitations in ERGM, Bayesian model applies Markov chain Monte Carlo (MCMC) simulation with the approximate exchange algorithm to estimate the double intractable posteriors [@Caimo2011]. This alleviate above computational problems. By fitting an ERMG with MCMC algorithm, a Bayesian approach provides posterior probability that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability enables us to estimate the effect of ERGM parameter by observing the posterior mean and 95% credible intervals, which can replace p-value for accessing a null hypothesis [@Caimo2017bayesian]. The criteria is that the odds ratio grater than one means positive effect of ERGM parameter, while odds ratio less than one represents negative effect.
